Solving the maximum independant set problem by generic programming einsum networks

### 2021-07-25 paper/paper.tex
update paper
clique counting

%\documentclass[aps,twocolumn,longbibliography,english,superscriptaddress]{revtex4-1}
\documentclass{article}
\usepackage{iclr2021_conference}
%\documentclass[a4paper,superscriptaddress,11pt]{article}
\pdfoutput=1
\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{quoting}
\usepackage{upquote}
\usepackage{subcaption}
\usepackage{multicol}
\usepackage[framemethod=TikZ]{mdframed}
\usetikzlibrary{shapes}
\usepackage{wrapfig}
%\usepackage{caption}
%\usepackage[plain]{algorithm}
\usepackage[linesnumbered, ruled, vlined]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{rotating}
%\usepackage{cite}
\usepackage{booktabs}
%\usepackage{unicode-math}
%\usepackage{algorithm}% http://ctan.org/pkg/algorithm
%\usepackage{algpseudocode}% http://ctan.org/pkg/algpseudocode
\usepackage{xcolor}% http://ctan.org/pkg/xcolor
\makeatletter
\newsavebox{\@brx}
\newcommand{\llangle}[1][]{\savebox{\@brx}{\(\m@th{#1\langle}\)}%
  \mathopen{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\newcommand{\rrangle}[1][]{\savebox{\@brx}{\(\m@th{#1\rangle}\)}%
  \mathclose{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\makeatother
\usepackage{bbm}
\usepackage{jlcode}
\usepackage{graphicx}
\usepackage{amsmath,color,amsthm}
\usepackage{mathrsfs}
\usepackage{float}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{indentfirst}
\usepackage{txfonts}
\usepackage[epsilon, tsrm, altpo]{backnaur}

\DeclareMathAlphabet{\mymathbb}{U}{BOONDOX-ds}{m}{n}
\newcommand{\listingcaption}[1]%
{%
\refstepcounter{lstlisting}\hfill%
Listing \thelstlisting: #1\hfill%\hfill%
}%
\newcolumntype{b}{X}
\newcolumntype{s}{>{\hsize=.7\hsize}X}
\usepackage{listings}
\lstset{
    language=Julia,
    basicstyle=\ttfamily\scriptsize,
    numberstyle=\scriptsize,
    % numbers=left,
    backgroundcolor=\color{gray!7},
    %backgroundcolor=\color{white},
    %frame=single,
    xleftmargin=2em,
    tabsize=2,
    rulecolor=\color{black!15},
    %title=\lstname,
    escapeinside={(*}{*)},
    breaklines=true,
    %breakatwhitespace=true,
    %framextopmargin=2pt,
    %framexbottommargin=2pt,
    frame=bt,
    extendedchars=true,
    inputencoding=utf8,
    columns=fullflexible,
    %escapeinside={(*@}{@*)},
}

@@

\iclrfinalcopy
\begin{document}
\title{Solving the maximum independant set problem by generic programming einsum networks}

\author{Jin-Guo Liu\\
Harvard University\\
\texttt{jinguoliu@g.harvard.edu}\\
\AND
Xun Gao\\
Harvard University\\
\texttt{xungao@g.harvard.edu}\\
}
\maketitle

@@

\section{Introduction}
Branching algorithms can solve the MIS problem in $1.1893^n n ^{O(1)}$ time ~\cite{Xiao2017}.
Previous dynamic programming approach~\cite{Fomin2013} can reduce the complexity of computing to $2^{tw(G)}$.

@@

One may also want to obtain all ground state configurations, it can be achieved replacing the factors $a_x$ with a set of bit strings $s_x$.
We design a new element type that having algebra
\iffalse
\begin{align}
\begin{split}
    s_x \oplus s_y &= s_x \cup s_y\\
    s_x \odot s_y &= \{\sigma \lor \tau | \sigma \in s_x, \tau \in s_y\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{\boldsymbol 0\}
\end{split}
\end{align}
\fi
\begin{align}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma + \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{\boldsymbol 0\}\infty^{0},
\end{split}
\end{align}
One can easily check that this replacement does not change the fact that the algebra is a commutative semiring.
We first initialize the bit strings of the variable $x$ in the vertex tensor to a vertex index $i$ dependent onehot vector $x_i = \boldsymbol{e}_{i}$,
then we contract the tensor network. The resulting object will give us the set of all optimal configurations.
By slightly modifying the above algebra, it can also be used to obtain just a single configuration to save computational effort.
We leave this as an exercise for readers.

@@

\section{Automated branching}
% check http://www.tcs.rwth-aachen.de/independentset/ for more rules
Branching rules can be automatically discovered by contracting the tropical einsum network for a subgraph $R \subseteq G$.
Let us denote the resulting tropical tensor of rank $|C|$ as $A$, where $C$ is the set of boundary vertices defined as $C := \{c | c\in R \land c \in G\backslash R\}$ and $|C|$ the size of $C$.
Each tensor entry $A_{\sigma}$ is a local maximum independant set size with a fixed boundary configuration $\sigma \in \{0,1\}^{|C|}$ by marginalizing the inner degrees of freedom.
If we are only interested in finding a single maximum independent set rather than enumerating all possible solutions,
this tensor can be further ``compresed'' by setting some entries to tropical zero.
Let us define a relation of \textit{less restrictive} as
\begin{align}
(\sigma_a \prec \sigma_b) := (\sigma_a \neq \sigma_b) \land (\sigma_a \leq^\circ \sigma_b)
\end{align}
where $\leq^\circ$ is the Hadamard less or equal operations.
We say an entry $A_{\sigma_a}$ is ``better'' than $A_{\sigma_b}$ if
\begin{align}
(\sigma_a \prec \sigma_b) \land (A_{\sigma_a} \geq A_{\sigma_b}),\label{eq:compress}.
\end{align}
If we remove such $A_{\sigma_b}$, the contraction over the whole graph is guaranted to give the same maximum independant set size.
It can be seen by considering two entries with the same local maximum independent set sizes and different boundary configurations as shown in \Fig{fig:compressrule} (a) and (b).
If we have $\sigma_b \cup \overline{\sigma_b}$ being one of the solutions for maximum independant sets in $G$, then $\sigma_a \cup \overline{\sigma_b}$ is another solution giving the same $\alpha(G)$.
Hence, we can set $A_{\sigma_b}$ to tropical zero safely.
%among boundary configurations with equal local maximum independent sizes,
%we only retain those least restrictive (less ones at the boundary) to exterial configurations.

@@

\begin{theorem}
    We denote a tensors $A$ as MIS-compact if are no two nonzero entries of it that one is ``better'' than another.
    A MIS-compact tropical tensor is optimal, i.e. none of its none zero entries can be removed without accessing global information.
\end{theorem}

@@

@@

Its einsum network representation is

\centerline{\begin{tikzpicture}
    \def\r{0.1}
    \def\a{0.07}
    \def\L{0.6}
    \def\l{0.1}
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
        \node[fill, regular polygon, 
            regular polygon sides=3, scale=0.3] at (\x, \y) {};
        }
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
            \ifnum \y < 5
                \filldraw[fill=black] (\x-\a,\y+0.5-\a) rectangle (\x+\a, \y+0.5+\a);
                \filldraw[fill=black] (\y+0.5-\a,\x-\a) rectangle (\y+0.5+\a, \x+\a);
            \fi
            \filldraw [red, opacity=0.3] plot [smooth cycle] coordinates {(\x-\L,\y+\l) (\x-\l,\y+\l) (\x-\l, \y+\L) (\x+\l,\y+\L) (\x+\l,\y+\l) (\x+\L,\y+\l) (\x+\L,\y-\l) (\x+\l,\y-\l) (\x+\l,\y-\L) (\x-\l,\y-\L) (\x-\l,\y-\l) (\x-\L,\y-\l)};
        }
    \foreach \x in {1,...,4}
        \foreach \y in {1,...,4}{
            \filldraw[fill=black] (\x+0.3-\a,\y+0.3-\a) rectangle (\x+0.3+\a, \y+0.3+\a);
            \filldraw[fill=black] (\y+0.3-\a,\x+0.7-\a) rectangle (\y+0.3+\a, \x+0.7+\a);
        }
\end{tikzpicture}}

>>>>

%\documentclass[aps,twocolumn,longbibliography,english,superscriptaddress]{revtex4-1}
\documentclass{article}
%\usepackage{iclr2021_conference}
\usepackage{natbib}
%\documentclass[a4paper,superscriptaddress,11pt]{article}
\pdfoutput=1
\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{quoting}
\usepackage{upquote}
\usepackage{subcaption}
\usepackage{multicol}
\usepackage[framemethod=TikZ]{mdframed}
\usetikzlibrary{shapes}
\usepackage{wrapfig}
%\usepackage{caption}
%\usepackage[plain]{algorithm}
\usepackage[linesnumbered, ruled, vlined]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{rotating}
%\usepackage{cite}
\usepackage{booktabs}
%\usepackage{unicode-math}
%\usepackage{algorithm}% http://ctan.org/pkg/algorithm
%\usepackage{algpseudocode}% http://ctan.org/pkg/algpseudocode
\usepackage{xcolor}% http://ctan.org/pkg/xcolor
\makeatletter
\newsavebox{\@brx}
\newcommand{\llangle}[1][]{\savebox{\@brx}{\(\m@th{#1\langle}\)}%
  \mathopen{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\newcommand{\rrangle}[1][]{\savebox{\@brx}{\(\m@th{#1\rangle}\)}%
  \mathclose{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\makeatother
\usepackage{bbm}
\usepackage{jlcode}
\usepackage{graphicx}
\usepackage{amsmath,color,amsthm}
\usepackage{mathrsfs}
\usepackage{float}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{indentfirst}
\usepackage{txfonts}
\usepackage[epsilon, tsrm, altpo]{backnaur}

\makeatletter
\def\parsept#1#2#3{%
    \def\nospace##1{\zap@space##1 \@empty}%
    \def\rawparsept(##1,##2){%
        \edef#1{\nospace{##1}}%
        \edef#2{\nospace{##2}}%
    }%
    \expandafter\rawparsept#3%
}
\makeatother
\DeclareMathAlphabet{\mymathbb}{U}{BOONDOX-ds}{m}{n}
\newcommand{\listingcaption}[1]%
{%
\refstepcounter{lstlisting}\hfill%
Listing \thelstlisting: #1\hfill%\hfill%
}%
\newcolumntype{b}{X}
\newcolumntype{s}{>{\hsize=.7\hsize}X}
\usepackage{listings}
\lstset{
    language=Julia,
    basicstyle=\ttfamily\scriptsize,
    numberstyle=\scriptsize,
    % numbers=left,
    backgroundcolor=\color{gray!7},
    %backgroundcolor=\color{white},
    %frame=single,
    xleftmargin=2em,
    tabsize=2,
    rulecolor=\color{black!15},
    %title=\lstname,
    escapeinside={(*}{*)},
    breaklines=true,
    %breakatwhitespace=true,
    %framextopmargin=2pt,
    %framexbottommargin=2pt,
    frame=bt,
    extendedchars=true,
    inputencoding=utf8,
    columns=fullflexible,
    %escapeinside={(*@}{@*)},
}

@@

%\iclrfinalcopy
\begin{document}
\title{Solving the maximum independant set problem by generic programming einsum networks}

%\author{Jin-Guo Liu\\
%Harvard University\\
%\texttt{jinguoliu@g.harvard.edu}\\
%\AND
%Xun Gao\\
%Harvard University\\
%\texttt{xungao@g.harvard.edu}\\
%}
\maketitle

@@

\section{Introduction}
MIS problem is hard~\cite{Hastad1996}.
Branching algorithms can solve the MIS problem in $1.1893^n n ^{O(1)}$ time ~\cite{Xiao2017}.
Previous dynamic programming approach~\cite{Fomin2013} can reduce the complexity of computing to $2^{tw(G)}$.
A set is independent if and only if it is a clique in the graph's complement, so the two concepts are complementary.
A set is independent if and only if its complement is a vertex cover.[4] Therefore, the sum of the size of the largest independent set $\alpha (G)$ and the size of a minimum vertex cover $\beta (G)$ is equal to the number of vertices in the graph.

@@

One may also want to obtain all ground state configurations, it can be achieved replacing the factors $a_x$ with a set of bit strings $s_x$.
We design a new element type that having algebra
\iffalse
\begin{align}
\begin{split}
    s_x \oplus s_y &= s_x \cup s_y\\
    s_x \odot s_y &= \{\sigma \lor \tau | \sigma \in s_x, \tau \in s_y\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{\boldsymbol 0\}
\end{split}
\end{align}
\fi
\begin{align}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma + \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{\boldsymbol 0\}\infty^{0},
\end{split}
\end{align}
One can easily check that this replacement does not change the fact that the algebra is a commutative semiring.
We first initialize the bit strings of the variable $x$ in the vertex tensor to a vertex index $i$ dependent onehot vector $x_i = \boldsymbol{e}_{i}$,
then we contract the tensor network. The resulting object will give us the set of all optimal configurations.
By slightly modifying the above algebra, it can also be used to obtain just a single configuration to save computational effort.
We leave this as an exercise for readers.
This algorithm is parallelizable.

@@

\section{Automated branching}
% check http://www.tcs.rwth-aachen.de/independentset/ for more rules
Branching rules can be automatically discovered by contracting the tropical einsum network for a subgraph $R \subseteq G$.
Let us denote the resulting tropical tensor of rank $|C|$ as $A$, where $C$ is the set of boundary vertices defined as $C := \{c | c\in R \land c \in G\backslash R\}$ and $|C|$ the size of $C$.
Each tensor entry $A_{\sigma}$ is a local maximum independant set size with a fixed boundary configuration $\sigma \in \{0,1\}^{|C|}$ by marginalizing the inner degrees of freedom.
If we are only interested in finding a single maximum independent set rather than enumerating all possible solutions,
this tensor can be further ``compresed'' by setting some entries to tropical zero.
Let us define a relation of \textit{less restrictive} as
\begin{align}
(\sigma_a \prec \sigma_b) := (\sigma_a \neq \sigma_b) \land (\sigma_a \leq^\circ \sigma_b)
\end{align}
where $\leq^\circ$ is the Hadamard less or equal operations.

\begin{definition}
A tensors $A$ is \textit{MIS-compact} if are no two nonzero entries of it that one is ``better'' than another,
where an entry $A_{\sigma_a}$ is ``better'' than $A_{\sigma_b}$ if
\begin{align}
(\sigma_a \prec \sigma_b) \land (A_{\sigma_a} \geq A_{\sigma_b})\label{eq:compress}.
\end{align}
\end{definition}

If we remove such $A_{\sigma_b}$, the contraction over the whole graph is guaranted to give the same maximum independant set size.
It can be seen by considering two entries with the same local maximum independent set sizes and different boundary configurations as shown in \Fig{fig:compressrule} (a) and (b).
If we have $\sigma_b \cup \overline{\sigma_b}$ being one of the solutions for maximum independant sets in $G$, then $\sigma_a \cup \overline{\sigma_b}$ is another solution giving the same $\alpha(G)$.
Hence, we can set $A_{\sigma_b}$ to tropical zero safely.
%among boundary configurations with equal local maximum independent sizes,
%we only retain those least restrictive (less ones at the boundary) to exterial configurations.

@@

\begin{theorem}
    A MIS-compact tropical tensor is optimal, i.e. none of its none zero entries can be removed without accessing global information.
\end{theorem}

@@

\subsection{gadget design}

Suppose we have a local structure as the following.

\centerline{\begin{tikzpicture}[scale=1.0]
    \def\r{0.2}
    \foreach \x/\y [count=\i] in {0.17/0.5,0.83/0.5,0.5/0.67,0.5/0.33}
        \node[fill=red,circle,radius=\r] at (\x*8, \y*8) (\i) {};
    \draw (1) -- (2);
    \draw (3) -- (4);
\end{tikzpicture}}

Contract this local structure gives the tropical tensor

\begin{align}
    \left(\begin{matrix}
        \left(\begin{matrix}
            ~~~~0 & ~~~~1\\
            ~~~~1 & ~~~~2
        \end{matrix}\right) &
        \left(\begin{matrix}
            ~~~~1 & -\infty\\
            ~~~~2 & -\infty
        \end{matrix}\right) \\
        \left(\begin{matrix}
            ~~~~1 & ~~~~2\\
            -\infty & -\infty
        \end{matrix}\right) &
        \left(\begin{matrix}
            ~~~~2 & -\infty\\
            -\infty & -\infty
        \end{matrix}\right)
    \end{matrix}\right).
\end{align}

The following gadget is equivalent to the above diagram up to a constant $2$.

\centerline{\begin{tikzpicture}
    \def\nodes{(0.5, 0.33), (0.17, 0.5), (0.5, 0.67), (0.83, 0.5), (0.32, 0.44), (0.44, 0.44), (0.56, 0.44), (0.68, 0.44), (0.32, 0.56), (0.44, 0.56), (0.56, 0.56), (0.68, 0.56)}
    \def\edges{(1, 5), (1, 6), (1, 7), (1, 8), (2, 5), (2, 9), (3, 9), (3, 10), (3, 11), (3, 12), (4, 8), (4, 12), (5, 6), (5, 9), (5, 10), (6, 7), (6, 9), (6, 10), (6, 11), (7, 8), (7, 10), (7, 11), (7, 12), (8, 11), (8, 12), (9, 10), (10, 11), (11, 12)}
    \def\r{0.2}
    \foreach \p [count=\i] in \nodes{
        \parsept{\x}{\y}{\p}
        \pgfmathparse{((\x<0.2 || \x>0.8 || \y<0.35 || \y>0.65) ? 1 : 0)}
        \ifnum\pgfmathresult>0
            \node[fill=red,circle,radius=\r] at (\x*8, \y*8) (\i) {};
        \else
            \node[fill,circle,radius=\r] at (\x*8, \y*8) (\i) {};
        \fi
    }
    \foreach \e [count=\i] in \edges{
        \parsept{\x}{\y}{\e}
        \draw (\x) -- (\y);
    }
\end{tikzpicture}}

\begin{align}
    \left(\begin{matrix}
        \left(\begin{matrix}
            ~~~~2 & ~~~~3\\
            ~~~~3 & ~~~~4
        \end{matrix}\right) &
        \left(\begin{matrix}
            ~~~~3 & ~~~~3\\
            ~~~~4 & ~~~~4
        \end{matrix}\right) \\
        \left(\begin{matrix}
            ~~~~3 & ~~~~4\\
            ~~~~2 & ~~~~3
        \end{matrix}\right) &
        \left(\begin{matrix}
            ~~~~4 & ~~~~4\\
            ~~~~3 & ~~~~4
        \end{matrix}\right)
    \end{matrix}\right)
    \xrightarrow[\text{compress, -2}]{}
    \left(\begin{matrix}
        \left(\begin{matrix}
            ~~~~0 & ~~~~1\\
            ~~~~1 & ~~~~2
        \end{matrix}\right) &
        \left(\begin{matrix}
            ~~~~1 & -\infty\\
            ~~~~2 & -\infty
        \end{matrix}\right) \\
        \left(\begin{matrix}
            ~~~~1 & ~~~~2\\
            -\infty & -\infty
        \end{matrix}\right) &
        \left(\begin{matrix}
            ~~~~2 & -\infty\\
            -\infty & -\infty
        \end{matrix}\right)
    \end{matrix}\right)
\end{align}

We can

\section{benchmarks}
We run a sequetial program benchmark on CPU Intel(R) Core(TM) i5-10400 CPU @ 2.90GHz, and show the results bellow.
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{benchmark.pdf}
    \caption{Benchmark results for computing different properties with different element types.
    }\label{fig:benchmark}
\end{figure}

@@

Once we represent a $\delta$ tensor as a general tensor, the complexity of this contraction is $\approx 2^{2L}$.
Its einsum network representation is

\centerline{\begin{tikzpicture}
    \def\r{0.1}
    \def\a{0.07}
    \def\L{0.6}
    \def\l{0.1}
    \def\sql{0.24}
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
        \node[fill, regular polygon, 
            regular polygon sides=3, scale=0.3] at (\x, \y) {};
        }
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
            \ifnum \y < 5
                \filldraw[fill=black] (\x-\a,\y+0.5-\a) rectangle (\x+\a, \y+0.5+\a);
                \filldraw[fill=black] (\y+0.5-\a,\x-\a) rectangle (\y+0.5+\a, \x+\a);
            \fi
            \filldraw [red, opacity=0.3] plot [smooth cycle] coordinates {(\x-\L,\y+\l) (\x-\l,\y+\l) (\x-\l, \y+\L) (\x+\l,\y+\L) (\x+\l,\y+\l) (\x+\L,\y+\l) (\x+\L,\y-\l) (\x+\l,\y-\l) (\x+\l,\y-\L) (\x-\l,\y-\L) (\x-\l,\y-\l) (\x-\L,\y-\l)};
        }
    \foreach \x in {1,...,4}
        \foreach \y in {1,...,4}{
            \filldraw[fill=black] (\x+0.3-\a,\y+0.3-\a) rectangle (\x+0.3+\a, \y+0.3+\a);
            \filldraw[fill=black] (\y+0.3-\a,\x+0.7-\a) rectangle (\y+0.3+\a, \x+0.7+\a);
        }
\end{tikzpicture}}
